Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-06-30 kappa.tex
Add introduction text.

@@

\begin{figure}
  \centering
  \includegraphics[width=80mm]{paleo-timeseries}
  \caption{Indicators of climatic fluctuations during the last 800\,ka of
           Earth history.
           \textbf{(a)} Benthic oxygen isotope \chem{\delta^{18}O} record from
           a compilation of globally-distributed oceanic sediment cores
           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
           correspond to periods of lower temperatures of the deep ocean, or
           of lower global sea level, and thus higher ice volume stored in
           glaciers and ice sheets on land.
           \textbf{(b)} Temperature anomalies reconstructed from a central
           Antarctic ice core \citep[EPICA,][]{Jouzel.etal.2007}.
           The grey boxes indicate the approximate time span of the last
           glacial period of 110 to 10\,ka, which corresponds to our period of
           interest in the present thesis.}
  \label{fig:palaeo-timeseries}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=80mm]{paleo-glaciation}
  \caption{Spatial extent of ice sheet and glacier cover in the northern
           hemisphere \textbf{(a)} at present, and \textbf{(b)} during the last
           glacial maximum \citep{Ehlers.Gibbard.2007}. Coastlines and present
           glaciers are made with Natural Earth \citep{Patterson.Kelso.2014}.}
  \label{fig:palaeo-glaciation}
\end{figure}

>>>>

In our daily lives, we most commonly experience ice as the small, solid cubes
that crackle on contact with refreshing liquids. Yet, at geological
time-scales, and under high stresses, ice behaves in a much different fashion.
In the cold and mountainous regions of the globe, ice accumulates under the
action of cumulative snowfall, until it reaches a critical thickness, and
deforms under its own weight to flow downhill. Glaciers, these bodies of moving
ice, flow by a combination of water-induced sliding at the base, and
visco-plastic deformation within the ice body (\cref{fig:glacier-mechanics},
\mref).

As ice flows and glide on its bed, it transports rock debris and erode the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive ice sheets had once covered much of Europe and North
America \mref.

However, the glacial theory did not gain much belief until the exploration and
discovery of the two present-day ice sheets on Earth, the Greenland and
Antarctic ice sheets, provided a modern analogue for the proposed European and
North American ice sheets. The first traverse of Greenland was performed by
Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
the European Alps, it was then discovered from observations of the glacial
landscape, that there had not been only one, but at least four, glacial periods
\mref.

Today, palaeo-climate records extracted from ice-core and oceanic sediment
records provide a much more detailed picture of the Earth environmental
history. Over the last few million years, temperature and global ice volume
have been constantly fluctuating, thus indicating that neither one, nor four,
but probably tens of glacial cycles have occured \mref. During the last million
years, glacial and interglacial cycles have succeeded each other with a 100-kyr
periodicity \mref. In the present thesis, we focus on the most recent of these
cycles: the last glacial cycle (\cref{fig:paleo-timeseries}).

Together, the Greenland and Antarctic ice sheets currently host more than two
thirds of the terrestrial fresh water, corresponding to 65~meters of potential
sea-level equivalent ice volume \mref. At the apogee of the last glacial cycle,
the last glacial maximum, sea-level was 120 to 135\,m lower than today \mref,
implying an additional, double amount of ice stored in ice sheets and glaciers
on land. Systematic investigation of glacial landforms on all continents
permitted reconstructions of this former ice cover, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasia arctic, and the
Innuitian, Laurentide and Cordilleran ice sheets over North America
(\cref{fig:paleo-glaciation}).

Nevertheless, the landscape record is timely sparse and spatially incomplete,
and geological reconstructions of former glacier extent demand a high amount of
interpretation. Palaeo-ice sheets did not leave a systematic record of their
passage, and much of the evidence has been overridden by multiple glacier
re-advance and other geomorphological processes. This is where numerical
models, incorporating the processes of glacier sliding and ice deformation
observed on modern glaciers, can help. Although these models alone contain too
many uncertainties to reproduce accurate picture of past glacial cycles, they
can be calibrated against field-based evidence where it exists, while allowing
a physics-based, quantitative reconstruction where it does not.

@@

\begin{figure}
  \centering
  \includegraphics[width=80mm]{paleo-timeseries}
  \caption{Indicators of climatic fluctuations during the last 800\,ka of
           Earth history.
           \textbf{(a)} Benthic oxygen isotope \chem{\delta^{18}O} record from
           a compilation of globally-distributed oceanic sediment cores
           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
           correspond to periods of lower temperatures of the deep ocean, or
           of lower global sea level, and thus higher ice volume stored in
           glaciers and ice sheets on land.
           \textbf{(b)} Temperature anomalies reconstructed from a central
           Antarctic ice core \citep[EPICA,][]{Jouzel.etal.2007}.
           The grey boxes indicate the approximate time span of the last
           glacial period of 110 to 10\,ka, which corresponds to our period of
           interest in the present thesis.}
  \label{fig:paleo-timeseries}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=80mm]{paleo-glaciation}
  \caption{Spatial extent of ice sheet and glacier cover in the northern
           hemisphere \textbf{(a)} at present, and \textbf{(b)} during the last
           glacial maximum \citep{Ehlers.Gibbard.2007}. Coastlines and present
           glaciers are made with Natural Earth \citep{Patterson.Kelso.2014}.}
  \label{fig:paleo-glaciation}
\end{figure}
